Shallow EDSLs and Object-Oriented Programming

### 2016-06-19 paper/sections/ShallowOO.tex
Draft Section 4

In this section, we show that OO approach and procedural abstraction are closely related
, and hence falls into the category of shallow embedding.
To demonstrate it, we use a subset
of the DSL presented in Gibbons and Wu's paper~\cite{gibbons2014folding} as the running example.
We first give the original shallow embedded implementation of the language
written in Haskell
and gradually rewrite it towards an ``OO style''.
Then translating the program into an OO language becomes straightforward.
We also illustrate how to do transformations in shallow embedding,

@@

\subsection{Shallow Embedding and Object-Oriented Programming}
Shallow embedding defines a language directly through encoding its semantics
using procedural abstraction. In the case of \dsl,
an shallow embedded implementation should conform to the following types:

\lstinputlisting[language=haskell,linerange=2-2]{./code/shallowCircuit.hs}%APPLY:CIRCUIT_TYPE
\lstinputlisting[language=haskell,linerange=5-7]{./code/shallowCircuit.hs}%APPLY:TYPES
The type \texttt{Circuit}, representing the semantic domain, is not defined yet
and should be filled in later with a concrete type according to the semantics.
Suppose that the semantics of \dsl is to calculate the width of a
circuit. The definitions would be:
\lstinputlisting[language=haskell,linerange=11-17]{./code/shallowCircuit.hs}%APPLY:CIRCUIT1
%As shallow embedding is semantics-oriented,
\texttt{width1} is an identity function because \texttt{Circuit} is a type
synonym of \lstinline{Int} and it is already an integer after circuit construction.
For example, we will immediately get $6$ for \texttt{beside (fan 3) (identity 3)}.

We start the rewrite by wrapping the result into an
datatype and getting back the value through pattern matching:
\lstinputlisting[language=haskell,linerange=21-28]{./code/shallowCircuit.hs}%APPLY:CIRCUIT2
% Then circuit construction gives back a \texttt{Circuit2} which contains an
% integer value representing the width and \texttt{width2} extracts out the value.
% Is this definition still shallow embedding?
%One may be curious whether this variant implementation is still shallow embedding or not. The answer is yes because ... % why it is shallow

The rewrite goes on by using a record with a sole field representing the result:
\lstinputlisting[language=haskell,linerange=33-38]{./code/shallowCircuit.hs}%APPLY:CIRCUIT3
We no longer need a separate \texttt{width3} function as the field name is the extractor.
The implementation is still shallow since essentially it is nothing different
from the last one.
However, the definition already approaches an OO program such that we can easily
translate it into an OO language like Scala, as shown in Figure~\ref{code:base}.

\begin{figure}
\lstinputlisting[linerange=2-13]{../src/scala/base/Circuit.scala}%APPLY:CIRCUIT_SCALA
\caption{OO implementation of \dsl}
\label{code:base}
\end{figure}
where the record type maps to the trait \texttt{Circuit} and field
declaration becomes method declaration.
Each case in the semantic function corresponds to an class and its parameters become fields of the class.
And all the classes extends \texttt{Circuit} by implementing \texttt{width}.

This implementation is exactly how we would model \dsl in an OO language in the first
place, following the \interp pattern (which uses \textsc{Composite} pattern to
organize classes).
In other words, shallow embedding and straightforward OO programming are closely
related.
It may worth mentioning that deep embedding is closely related to the \textsc{Visitor} pattern.

>>>>

In this section, we show that OO approach and procedural abstraction are closely
related, and hence falls into the category of shallow embedding.
To demonstrate it, we use a subset
of the DSL presented in Gibbons and Wu's paper~\cite{gibbons2014folding} as the running example.
We first give the original shallow embedded implementation in Haskell
and gradually rewrite it towards an ``OO style''.
Then translating the program into an OO language becomes straightforward.

@@

\subsection{Shallow Embedding and OOP}
Shallow embedding defines a language directly through encoding its semantics
using procedural abstraction. In the case of \dsl,
an shallow embedded implementation should conform to the following types:

\lstinputlisting[language=haskell,linerange=2-2]{./code/shallowCircuit.hs}%APPLY:CIRCUIT_TYPE
\lstinputlisting[language=haskell,linerange=5-7]{./code/shallowCircuit.hs}%APPLY:TYPES
The type \lstinline{Circuit}, representing the semantic domain, is to be filled in with a concrete type according to the semantics.
Suppose that the semantics of \dsl is to calculate the width of a
circuit. The definitions would be:
\lstinputlisting[language=haskell,linerange=11-17]{./code/shallowCircuit.hs}%APPLY:CIRCUIT1
%As shallow embedding is semantics-oriented,
\lstinline{width1} is an identity function because \lstinline{Circuit} is a type
synonym of \lstinline{Int} and it is already an integer after circuit construction.
For example, we will immediately get $6$ for \lstinline{beside (fan 3) (identity 3)}.

We start the rewrite by wrapping the result into an
datatype and getting back the value through pattern matching:
\lstinputlisting[language=haskell,linerange=21-28]{./code/shallowCircuit.hs}%APPLY:CIRCUIT2
% Then circuit construction gives back a \lstinline{Circuit2} which contains an
% integer value representing the width and \lstinline{width2} extracts out the value.
% Is this definition still shallow embedding?
%One may be curious whether this variant implementation is still shallow embedding or not. The answer is yes because ... % why it is shallow

The rewrite goes on by using a record with a sole field representing the result:
\lstinputlisting[language=haskell,linerange=33-38]{./code/shallowCircuit.hs}%APPLY:CIRCUIT3
We no longer need a separate \lstinline{width3} function as the field name is the extractor.
The implementation is still shallow since essentially it is nothing different
from the last one.
However, the definition already approaches an OO program such that we can easily
translate it into an OO language like Scala, as shown in Figure~\ref{code:base}.

\begin{figure}
\lstinputlisting[linerange=2-13]{../src/scala/base/Circuit.scala}%APPLY:CIRCUIT_SCALA
\caption{OO implementation of \dsl}
\label{code:base}
\end{figure}
where the record type maps to the trait \lstinline{Circuit} and field
declaration becomes method declaration.
Each case in the semantic function corresponds to an class and its parameters become fields of the class.
And all the classes extends \lstinline{Circuit} by implementing \lstinline{width}.

This implementation is exactly how we would model \dsl in an OO language in the first
place, following the \interp pattern (which uses \textsc{Composite} pattern to
organize classes).
In summary, shallow embedding and straightforward OO programming are closely
related.
It may worth mentioning that deep embedding is closely related to the \textsc{Visitor} pattern.

### 2016-06-19 paper/sections/Transformations.tex
Draft Section 4

Figure~\ref{code:desugar} shows how to implement desugaring in Scala.
\begin{figure}
\lstinputlisting[linerange=3-18]{../src/scala/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
\caption{OO implementation of desugaring}
\label{code:desugar}
\end{figure}
The newly introduced method \texttt{desugar} in \texttt{Circuit} returns a \texttt{Circuit} reflecting that it is a transformation. The implementation of \texttt{desugar} for each case is
straightforward: \emph{fan n} just returns itself; $beside\ c_1\ c_2$ recursively desugars $c_1$ and $c_2$ then wraps back
their results into an instance of \texttt{Beside}; \emph{identity n} mimics the
formula.

%\weixin{define a show function in the and show the result after desugaring?}
Inspired by the Scala implementation, we can also define \texttt{desugar} in Haskell:

\lstinputlisting[language=haskell,linerange=6-21]{./code/Desugar.hs}%APPLY:DESUGAR
The definition relies on the laziness of Haskell.
However, different from the Scala version, introducing \texttt{desugar} is done
through modifying the original code by inserting a new field in the record.
We will discuss modularity issues in detail in Section~\ref{}.

@@

\paragraph{Extractors.}
Scala supports pattern matching.
one can manually implement an extractor, a.k.a \texttt{unapply}
method, in the companion object to tell how to destruct an object.
\begin{lstlisting}
object Id {
  def unapply(c: Identity) = Some(c.n)
}

>>>>

Figure~\ref{code:desugar} shows how to implement desugaring in Scala.
\begin{figure}
\lstinputlisting[linerange=3-18]{../src/scala/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
\caption{OO implementation of desugaring}
\label{code:desugar}
\end{figure}
The newly introduced method \lstinline{desugar} in \lstinline{Circuit} returns a \lstinline{Circuit} reflecting that it is a transformation. The implementation of \lstinline{desugar} for each case is
straightforward: \emph{fan n} just returns itself; $beside\ c_1\ c_2$ recursively desugars $c_1$ and $c_2$ then wraps back
their results into an instance of \lstinline{Beside}; \emph{identity n} mimics the
formula.

%\weixin{define a show function in the and show the result after desugaring?}
Inspired by the Scala implementation, we can also define \lstinline{desugar} in Haskell:

\begin{figure}
\lstinputlisting[language=haskell,linerange=6-21]{./code/Desugar.hs}%APPLY:DESUGAR
\caption{Haskell implementation of desugaring}
\label{code:haskell}
\end{figure}
The definition relies on the laziness of Haskell.
However, different from the Scala version, introducing \lstinline{desugar} is done
through modifying the original code by inserting a new field in the record.
We will discuss modularity issues in detail in Section~\ref{}.

@@

\paragraph{Extractors.}
Scala supports pattern matching.
one can manually implement an extractor, a.k.a \lstinline{unapply}
method, in the companion object to tell how to destruct an object.
\begin{lstlisting}
object Id {
  def unapply(c: Identity) = Some(c.n)
}
